

%%% Choose a language %%%

\newif\ifEN
\ENtrue   % uncomment this for english
%\ENfalse   % uncomment this for czech

%%% Configuration of the title page %%%

\newif\ifMFF
\MFFtrue % comment this out for the version with a big UK university logo
\def\UKName{Charles University in Prague} %this is used in UK-logo-version
\def\UKFaculty{Faculty of Mathematics and Physics}
\def\ThesisTypeName{\ifEN BACHELOR THESIS \else BAKALÁŘSKÁ PRÁCE \fi}
%\def\ThesisTypeName{\ifEN MASTER THESIS \else DIPLOMOVÁ PRÁCE \fi}
%\def\ThesisTypeName{\ifEN RIGOROUS THESIS \else RIGORÓZNÍ PRÁCE \fi}
%\def\ThesisTypeName{\ifEN DOCTORAL THESIS \else DISERTAČNÍ PRÁCE \fi}
\def\SupervisedThesisName{\ifEN bachelor \else bakalářské \fi}
%\def\SupervisedThesisName{\ifEN master \else diplomové \fi}
%\def\SupervisedThesisName{\ifEN rigorous \else rigorózní \fi}
%\def\SupervisedThesisName{\ifEN doctoral \else disertační \fi}


%%% Fill in your details %%%

% (Note: \xxx is a "ToDo label" which makes the unfilled visible. Remove it.)
\def\ThesisTitle{Type Inference and Polymorphism for C}
\def\ThesisAuthor{Jiří Klepl}
\def\YearSubmitted{2020}

% department assigned to the thesis
\def\Department{Department of Software Engineering}
% Is it a department (katedra), or an institute (ústav)?
\def\DeptType{Department}

\def\Supervisor{RNDr. Miroslav Kratochvíl}
\def\SupervisorsDepartment{Department of Software Engineering}

% Study programme and specialization
\def\StudyProgramme{Computer Science}
\def\StudyBranch{General Computer Science}

\def\Dedication{%
I would like to thank my supervisor, RNDr. Miroslav Kratochvíl, for his guidance and active support during my work on this thesis, and for introducing me to functional programming.
}

\def\Abstract{%
The C languages, despite its age, is one of the main languages in systems development. It is valued for giving the user almost complete control over the memory management and the computations a program written in it performs.

A large portion of criticism of C arises from the lack of generic programming features. C compensates that by macros, which are prone to user errors.

This problem has been addressed in the early stages of the C++ language, but many systems developers refuse C++ because of its complexity and non-transparency of the code.

We propose a simpler solution by applying the Hindley-Milner type system extended by Haskell type classes and type constructors. We will show that this approach is viable even with minimal changes to C, giving it much higher expressiveness.
}

% 3 to 5 keywords (recommended), each enclosed in curly braces.
% Keywords are useful for indexing and searching for the theses by topic.
\def\Keywords{%
{C} {Hindley} {Milner} {Polymorphism} {Type-Inference}
}

% If your abstracts are long and do not fit in the info-page, you can make the
% fonts a bit smaller by this setting. (Also, you should try to compress your abstract more.)
% Alternatively, consider increasing the size of the page by uncommenting the
% geometry modification in thesis.tex.
\def\InfoPageFont{}
%\def\InfoPageFont{\small}  %uncomment to decrease font size

\ifEN\relax\else
% If you are writing a czech thesis, you additionally need to fill in the
% english translation of the metadata here!
\def\ThesisTitleEN{Type Inference and Polymorphism for C}
\def\DepartmentEN{\xxx{Name of the department in English}}
\def\DeptTypeEN{\xxx{Department}}
\def\SupervisorsDepartmentEN{\xxx{Superdepartment}}
\def\StudyProgrammeEN{\xxx{study programme}}
\def\StudyBranchEN{\xxx{study branch}}
\def\AbstractEN{%
\xxx{Abstract.}
}
\def\KeywordsEN{%
\xxx{{key} {words}}
}
\fi
